The Changing Shape of Empire: A GM Study of Chimú Bottles in Museum Collections

### 2022-02-04 supp/05-gm.body.Rmd

<<<<<<< HEAD:supp/04-gm.sculptural.Rmd
# Sculptural elements + Vessel form
=======
# 3DGM: Neck morphology
>>>>>>> 61d5dbccf7b8bd18270ef346337ac517016a6843:supp/04-gm.neck.Rmd

@@

```{r load.data.s, echo = TRUE}
# load packages
#devtools::install_github("mlcollyer/RRPP")
#devtools::install_github("geomorphR/geomorph", ref = "Stable", build_vignettes = TRUE)
library(here)
library(geomorph)
library(tidyverse)
library(wesanderson)

@@

```{r gpa.s, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
#select landmarks associated with neck
a <- c(1:10, 36:45)

# new coords neck
neck.coords <- coords[a,,]

# gpa
Y.gpa <- gpagen(neck.coords, 
                PrinAxes = TRUE, 
                ProcD = TRUE, 
                Proj = TRUE, 
                print.progress = FALSE)

@@

# render 3d gpa plot
plot(Y.gpa)

@@

```{r box1.s, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, fig.cap = "Boxplot of centroid size by spatial/temporal unit."}
# attributes for boxplot
csz <- qdata$csz
type <-  qdata$type

@@

# boxplot of vessel centroid size by type
csz.type <- ggplot(qdata, aes(x = type, y = csz, color = type)) + 
  geom_boxplot(notch = FALSE) +
  geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 0.3) +
  scale_colour_manual(values = pal) +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(x = 'Type (Neck)', 
       y = 'Centroid Size')

@@

```{r gmpca.s, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# principal components analysis
pca<-gm.prcomp(Y.gpa$coords)
summary(pca)

@@

```{r pca.warp.s, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, fig.cap="Results of PCA summarising shape variation in the aggregated sample; gray, north-formative/early; orange, north-late/historic; brown, south-formative/early; and black, south-late/historic; with mean shapes (gray) contrasted with max/min shapes (black)."}
# pca warp                   
#knitr::include_graphics('images/pca-warp.jpg')
```

## Neck size and shape

@@

```{r def.mod.1.s, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# neck size as a function of vessel form (bottle/jar)?
fit.size.type <- procD.lm(size ~ type,
                            data = gdf,
                            print.progress = FALSE,
                            iter = 9999)

# Procrustes ANOVA: do neck sizes differ as a function of vessel form?
anova(fit.size.type)

# neck shape as a function of vessel form?
fit.shape.type <- procD.lm(shape ~ type,
                data = gdf,
                print.progress = FALSE,
                iter = 9999)

# Procrustes ANOVA: do neck shapes differ as a function of vessel form?
anova(fit.shape.type)
```

@@

```{r mshape.s, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# subset landmark coordinates to produce mean shapes
new.coords<-coords.subset(A = Y.gpa$coords,
                          group = qdata$type)
names(new.coords)

@@

knitr::include_graphics('images/mshape-neck.png')
```

>>>>

# 3DGM: Body morphology

@@

```{r load.data.b, echo = TRUE}
# load packages
#devtools::install_github("mlcollyer/RRPP")
#devtools::install_github("geomorphR/geomorph", ref = "Stable", build_vignettes = TRUE)
library(here)
library(geomorph)
library(tidyverse)
library(wesanderson)

@@

```{r gpa.b, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
#select landmarks associated with body
b <- c(10:19, 27:36)

# new coords body
body.coords <- coords[b,,]

# gpa
Y.gpa <- gpagen(body.coords, 
                PrinAxes = TRUE, 
                ProcD = TRUE, 
                Proj = TRUE, 
                print.progress = FALSE)

@@

# render 3d gpa plot
#plot(Y.gpa)

@@

```{r box1.b, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, fig.cap = "Boxplot of centroid size by spatial/temporal unit."}
# attributes for boxplot
csz <- qdata$csz
type <-  qdata$type

@@

# boxplot of vessel centroid size by type
csz.type <- ggplot(qdata, aes(x = type, y = csz, color = type)) + 
  geom_boxplot(notch = FALSE) +
  geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 0.3) +
  scale_colour_manual(values = pal) +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(x = 'Type (Body)', 
       y = 'Centroid Size')

@@

```{r gmpca.b, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# principal components analysis
pca<-gm.prcomp(Y.gpa$coords)
summary(pca)

@@

```{r pca.warp.b, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, fig.cap="Results of PCA summarising shape variation in the aggregated sample; gray, north-formative/early; orange, north-late/historic; brown, south-formative/early; and black, south-late/historic; with mean shapes (gray) contrasted with max/min shapes (black)."}
# pca warp                   
#knitr::include_graphics('images/pca-warp.jpg')
```

## Body size and shape

@@

```{r def.mod.1.b, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# body size as a function of vessel form (bottle/jar)?
fit.size.type <- procD.lm(size ~ type,
                            data = gdf,
                            print.progress = FALSE,
                            iter = 9999)

# Procrustes ANOVA: do body sizes differ as a function of vessel form?
anova(fit.size.type)

# body shape as a function of vessel form?
fit.shape.type <- procD.lm(shape ~ type,
                data = gdf,
                print.progress = FALSE,
                iter = 9999)

# Procrustes ANOVA: do body shapes differ as a function of vessel form?
anova(fit.shape.type)
```

@@

```{r mshape.b, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# subset landmark coordinates to produce mean shapes
new.coords<-coords.subset(A = Y.gpa$coords,
                          group = qdata$type)
names(new.coords)

@@

#knitr::include_graphics('images/mshape-body.png')
```

### 2022-02-04 supp/06-gm.base.Rmd

```{r load.data.b, echo = TRUE}
# load packages
#devtools::install_github("mlcollyer/RRPP")
#devtools::install_github("geomorphR/geomorph", ref = "Stable", build_vignettes = TRUE)
library(here)
library(geomorph)
library(tidyverse)
library(wesanderson)

@@

```{r gpa.b, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
#select landmarks associated with body
b <- c(10:19, 27:36)

# new coords body
body.coords <- coords[b,,]

# gpa
Y.gpa <- gpagen(body.coords, 
                PrinAxes = TRUE, 
                ProcD = TRUE, 
                Proj = TRUE, 
                print.progress = FALSE)

@@

```{r box1.b, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, fig.cap = "Boxplot of centroid size by spatial/temporal unit."}
# attributes for boxplot
csz <- qdata$csz
type <-  qdata$type

@@

# boxplot of vessel centroid size by type
csz.type <- ggplot(qdata, aes(x = type, y = csz, color = type)) + 
  geom_boxplot(notch = FALSE) +
  geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 0.3) +
  scale_colour_manual(values = pal) +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(x = 'Type (Body)', 
       y = 'Centroid Size')

@@

```{r gmpca.b, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# principal components analysis
pca<-gm.prcomp(Y.gpa$coords)
summary(pca)

@@

```{r pca.warp.b, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, fig.cap="Results of PCA summarising shape variation in the aggregated sample; gray, north-formative/early; orange, north-late/historic; brown, south-formative/early; and black, south-late/historic; with mean shapes (gray) contrasted with max/min shapes (black)."}
# pca warp                   
#knitr::include_graphics('images/pca-warp.jpg')
```

@@

```{r def.mod.1.b, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# body size as a function of vessel form (bottle/jar)?
fit.size.type <- procD.lm(size ~ type,
                            data = gdf,
                            print.progress = FALSE,
                            iter = 9999)

@@

```{r mshape.b, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# subset landmark coordinates to produce mean shapes
new.coords<-coords.subset(A = Y.gpa$coords,
                          group = qdata$type)
names(new.coords)

@@

#knitr::include_graphics('images/mshape-body.png')
```

>>>>

```{r load.data.ba, echo = TRUE}
# load packages
#devtools::install_github("mlcollyer/RRPP")
#devtools::install_github("geomorphR/geomorph", ref = "Stable", build_vignettes = TRUE)
library(here)
library(geomorph)
library(tidyverse)
library(wesanderson)

@@

```{r gpa.ba, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
#select landmarks associated with base
ba <- c(19:27)

# new coords base
base.coords <- coords[ba,,]

# gpa
Y.gpa <- gpagen(base.coords, 
                PrinAxes = TRUE, 
                ProcD = TRUE, 
                Proj = TRUE, 
                print.progress = FALSE)

@@

```{r box1.ba, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, fig.cap = "Boxplot of centroid size by spatial/temporal unit."}
# attributes for boxplot
csz <- qdata$csz
type <-  qdata$type

@@

# boxplot of vessel centroid size by type
csz.type <- ggplot(qdata, aes(x = type, y = csz, color = type)) + 
  geom_boxplot(notch = FALSE) +
  geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 0.3) +
  scale_colour_manual(values = pal) +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(x = 'Type (Base)', 
       y = 'Centroid Size')

@@

```{r gmpca.ba, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# principal components analysis
pca<-gm.prcomp(Y.gpa$coords)
summary(pca)

@@

```{r pca.warp.ba, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, fig.cap="Results of PCA summarising shape variation in the aggregated sample; gray, north-formative/early; orange, north-late/historic; brown, south-formative/early; and black, south-late/historic; with mean shapes (gray) contrasted with max/min shapes (black)."}
# pca warp                   
#knitr::include_graphics('images/pca-warp.jpg')
```

@@

```{r def.mod.1.ba, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# body size as a function of vessel form (bottle/jar)?
fit.size.type <- procD.lm(size ~ type,
                            data = gdf,
                            print.progress = FALSE,
                            iter = 9999)

@@

```{r mshape.ba, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# subset landmark coordinates to produce mean shapes
new.coords<-coords.subset(A = Y.gpa$coords,
                          group = qdata$type)
names(new.coords)

@@

#knitr::include_graphics('images/mshape-base.png')
```

### 2022-02-04 supp/index.Rmd

### Geometric morphometrics

@@

>>>>

### 3D Geometric morphometrics

**Whole vessel morphology**:

@@

**Neck morphology**:

 * Bottle and jar necks **differ significantly** in shape
 * Bottle and jar necks do not differ in size

```{r botjarneck, out.width = "100%", dpi = 300, echo=FALSE, warning=FALSE, fig.cap="Mean shapes for bottle necks (a), jar necks (b), and comparison of bottle (gray) and jar (black) neck shapes (c)."}
knitr::include_graphics('images/mshape-neck.png')
```

**Body morphology**:

 * Bottle and jar bodies **differ significantly** in shape
 * Bottle and jar bodies do not differ in size

**Base morphology**:

 * Bottle and jar bases **differ significantly** in shape
 * Bottle and jar bases do not differ in size
